# CART genome integrations

Here we will plot where CAR T constructs are integrated in the genome.

```{r set up, warning = FALSE, message = FALSE}
# Source setup file including all required packages and functions
source("./functions/setup.R")
```

### Get CART supplementary reads alignments

First we run the following bash scripts to get all the primary (CAR construct) and supplementary (integration site) alignment locations.

```{bash get CART reads, eval = FALSE}
samtools view -F 0x400 ./data/scdnalong/results_ref_carT_contig_deduped/Q3/bam/dedup/Q3.dedup.sorted.bam CTL019_6k | \
awk 'BEGIN{OFS="\t"} !/^@/ {
    if (match($0, /SA:Z:([^,]+),([0-9]+),/, m)) {
        print $1, $3, $4, m[1], m[2];
    }
}' > ./data/CART/supplreads_coordinates_untargeted.tsv
```


### Load in data

First load in the previously generated coordinate file.

```{r load in data}
dt = fread("./data/CART/supplreads_coordinates_untargeted.tsv")
cytobands = read.cytoband(species = "hg38")
```

### Preprocess data

Then prepare data for plotting

```{r preprocess data}
# Remove rows where supplementary is same contig
dt = dt[V2 != V4, ]
dt = dt[!grepl("decoy", V4) & !grepl("chrUn", V4), ]


# Prepare dataframes
primary = dt[, .(chr = V2, start = V3, end = V3, sample = V1)]
supplementary = dt[, .(chr = V4, start = V5, end = V5, sample = V1)]

chroms = unique(c(primary$chr, supplementary$chr))
chroms = gtools::mixedsort(chroms)

# Prepare cytobands
setDT(cytobands$df)
cytobands_dt = cytobands$df[V1 %in% chroms, ]

CAR_Contig = data.table(V1 = "CTL019_6k", V2 = 0, V3 = 5406, V4 = "CTL019", V5 = "gneg")

final_cytobands = rbind(cytobands_dt, CAR_Contig)

# Make custom sector width to increase CTL019_6k sector size
xrange = c(cytobands$chr.len, CTL019_6k = final_cytobands[V1 == "CTL019_6k", V3])
xrange = xrange[chroms]

normal_chr_index = seq_along(xrange) - 1
zoomed_chr_index = length(xrange)
sector.width = c(xrange[normal_chr_index] / sum(xrange[normal_chr_index]),
								 xrange[zoomed_chr_index] / sum(xrange[zoomed_chr_index]) / 14)
```

### Plot UMAP

Finally we plot the UMAP with CART status.

```{r plot circos}
circos.par("start.degree" = 90)
circos.initializeWithIdeogram(final_cytobands, sector.width = sector.width, chromosome.index = chroms)
circos.genomicLink(primary, supplementary)
```