# Isoform quantification

Here we will plot the isoform quantification results from the combined analysis of all timepoints.

First we load packages. These need to be preinstalled in your R environment. You can find all packages in the `setup.R` file in the `functions` folder.
Furthermore, we also load functions that help us with plotting and saving.


### Merging and collapsing of transcipts

Merging is done using TAMA merge, which is part of the TAMA tool suite. We use the default settings for merging.

```{bash merge isoforms}
python tama_merge.py \
    -f ${mergefile} \
    -p ${outdir}/merged

```

Then we collapse similar transcripts using TAMA collapse, which is also part of the TAMA tool suite. We use the default settings for collapsing, which means that all transcripts that have the same splice junctions and are within 100 bp of each other will be collapsed together. The only thing we specify is the output prefix and the input file, which is the merged file from the previous step.

```{bash collapse isoforms}
python tama_collapse.py \
    -b BAM \
    -s ${bam} \
    -f ${ref} \
    -p ${outdir}/B_ALL \
    -x no_cap 
```


### QC

After obtaining the collapsed transcriptome, we can run SQANTI3 to get QC metrics on the isoform quantification and perform filtering, only retaining high confidence isoform calls.
For filtering we used an adapted filtering scheme which you can see below.

```{json adapted filter settings}
{
    "full-splice_match": [
        {
            "perc_A_downstream_TTS":[0,59]
        }
    ],
    "rest": [
        {
            "perc_A_downstream_TTS":[0,59],
            "all_canonical":"canonical",
            "RTS_stage":"FALSE"
        }
     ]
}

```

Then we created a `yaml` config file for SQANTI3 with the appropriate paths and settings, which is used to run SQANTI3.

An example of the yaml file:
```{yaml example yaml}
main:
  refGTF: '/references/10X_CellRanger_refdata-gex-GRCh38-2020-A/genes/genes.gtf'
  refFasta: '/references/10X_CellRanger_refdata-gex-GRCh38-2020-A/fasta/genome.fa'
  cpus: 8
  dir: ./sqanti3_default_adapted
  output: isoforms
  log_level: INFO
qc:
  enabled: true
  options:
    isoforms: './tama/merged.gtf'
    min_ref_len: 0
    force_id_ignore: false
    fasta: false
    genename: false
    novel_gene_prefix: ''
    sites: ATAC,GCAG,GTAG
    window: 20
    aligner_choice: minimap2
    gmap_index: ''
    skipORF: false
    orf_input: ''
    short_reads: ''
    SR_bam: ''
    CAGE_peak: ''
    polyA_motif_list: ''
    polyA_peak: ''
    phyloP_bed: ''
    expression: ''
    coverage: ''
    fl_count: ''
    isoAnnotLite: false
    gff3: ''
    saturation: false
    report: html
    isoform_hits: false
    ratio_TSS_metric: max
    chunks: 1
    is_fusion: false
    tusco: ''
filter:
  enabled: true
  options:
    common:
      sqanti_class: ./sqanti3_default_adapted/isoforms_classification.txt
      isoAnnotGFF3: ''
      filter_isoforms: ./sqanti3_default_adapted/isoforms_corrected.fasta
      filter_gtf: ./sqanti3_default_adapted/isoforms_corrected.gtf
      filter_sam: ''
      filter_faa: ./sqanti3_default_adapted/isoforms_corrected.faa
      skip_report: false
      filter_mono_exonic: false
    rules:
      enabled: true
      options:
        json_filter: ./filter_default_adapted.json
    ml:
      enabled: false
      options:
        percent_training: 0.8
        TP: ''
        TN: ''
        threshold: 0.7
        force_fsm_in: false
        intermediate_files: false
        remove_columns: ''
        max_class_size: 3000
        intrapriming: 60
rescue:
  enabled: true
  options:
    common:
      filter_class: ./sqanti3_default_adapted/isoforms_RulesFilter_result_classification.txt
      rescue_isoforms: ./sqanti3_default_adapted/isoforms_corrected.fasta
      rescue_gtf: ./sqanti3_default_adapted/isoforms.filtered.gtf
      refClassif: './sqanti3_default_adapted/isoforms_classification.txt'
      counts: ''
      rescue_mono_exonic: all
      mode: automatic
      requant: false
      random_forest: ''
    rules:
      enabled: true
      options:
        json_filter: ./filter_default_adapted.json
    ml:
      enabled: false
      options:
        random_forest: ''
        threshold: 0.7
```

SQANTI3 is then run as follows:

```{bash SQANTI3 filtering}
sqanti3 all -c "config.yaml" \
		-a isoforms="${gtf_file}" \
		-a dir="${sqanti_dir}" \
		-a refGTF="${ref_gtf}" \
		-a refFasta="${ref_fasta}" \
		-a output="merged" \
		-a cpus="${SLURM_CPUS_PER_TASK}"
```


```{r set up, warning = FALSE, message = FALSE}
# Source setup file including all required packages and functions
source("./functions/setup.R")
```


Then we plot the number of isoforms detected after filtering.
Since the numbers are saved in a pdf we did this manually

```{r plot isoform quantification unique}
dt = data.table(type = c("FSM", "ISM", "NIC", "NNC", "Genic", "Antisense", "Genic fusion", "Intergenic", "Genic intron"),
				count = c(2262, 606, 75, 224, 42, 488, 2, 150, 263))
dt[, type := factor(type, levels = c("FSM", "ISM", "NIC", "NNC", "Genic", "Antisense", "Genic fusion", "Intergenic", "Genic intron"))]

# Plot
plt = ggplot(dt, aes(x = type, y = count, group = type, fill = type)) +
	geom_col(position = "dodge") +
	scale_fill_brewer(palette = "Paired") +
	theme_cowplot() +
	labs(y = "Number of transcripts", x = "") +
	scale_y_continuous(expand = c(0, 0), labels = scales::comma_format()) +
	theme(legend.position = "none",
				text = element_text(family = "Helvetica"),
				axis.text.x = element_text(angle = 45, hjust = 1))
```