# Isoform quantification

Here we will plot the isoform quantification results from the combined analysis of all timepoints.

First we load packages. These need to be preinstalled in your R environment. You can find all packages in the `setup.R` file in the `functions` folder.
Furthermore, we also load functions that help us with plotting and saving.

```{r set up, warning = FALSE, message = FALSE}
# Source setup file including all required packages and functions
source("./functions/setup.R")
```


### Load in data

```{r load in data}
# Load in transcripts
combined_counts = fread("./data/combined_grouped_transcript_model_counts.tsv")
cells_doublets = fread("./data/barcode_doublet.tsv")
```

### Filter out doublets

First we filter out doublet cells that were manually identified

```{r filter out doublets}
# Restructure cell barcode in doublets
cells_doublets[, timepoint := gsub(".*-", "", V1)]
cells_doublets[, barcode := gsub("-.*", "", V1)]
cells_doublets[, clean_barcode := paste0(timepoint, "_", barcode)]

# Rename
setnames(combined_counts, "#feature_id", "transcript_id")

# Filter out doublets
combined_counts_filtered = combined_counts[, !colnames(combined_counts) %in% cells_doublets[doublet_outlier == TRUE, clean_barcode], with = FALSE]
```


### Preprocess data

We split the data per timepoint and get the sum of each transcript per timepoint.

```{r preprocess data}
# Split per timepoint
counts_d0 = combined_counts[, c("transcript_id", grep("D0", colnames(combined_counts), value = TRUE)), with = FALSE]
counts_q1 = combined_counts[, c("transcript_id", grep("Q1", colnames(combined_counts), value = TRUE)), with = FALSE]
counts_q2 = combined_counts[, c("transcript_id", grep("Q2", colnames(combined_counts), value = TRUE)), with = FALSE]
counts_q3 = combined_counts[, c("transcript_id", grep("Q3", colnames(combined_counts), value = TRUE)), with = FALSE]

# Get sum of each transcript per timepoint
counts_d0[, sum := rowSums(.SD), .SDcols = -1]
counts_q1[, sum := rowSums(.SD), .SDcols = -1]
counts_q2[, sum := rowSums(.SD), .SDcols = -1]
counts_q3[, sum := rowSums(.SD), .SDcols = -1]

# Get counts per transcript type per timepoint
counts_d0[, type := ifelse(grepl("ENST", transcript_id), "known", gsub(".*\\.", "", transcript_id))]
counts_q1[, type := ifelse(grepl("ENST", transcript_id), "known", gsub(".*\\.", "", transcript_id))]
counts_q2[, type := ifelse(grepl("ENST", transcript_id), "known", gsub(".*\\.", "", transcript_id))]
counts_q3[, type := ifelse(grepl("ENST", transcript_id), "known", gsub(".*\\.", "", transcript_id))]

# Make new data.table
dt = data.table(transcript_id = counts_d0$transcript_id,
								D0 = counts_d0$sum,
								Q1 = counts_q1$sum,
								Q2 = counts_q2$sum,
								Q3 = counts_q3$sum,
								type = counts_d0$type)
dt_m = melt(dt, id.vars = c("transcript_id", "type"), variable.name = "timepoint", value.name = "count")

dt_m_aggr = dt_m[, .(count = sum(count)), by = c("type", "timepoint")]
dt_m_aggr[, type := factor(type, levels = c("known", "nnic", "nic"))]
```

### Plot isoform quantification

First we plot the total number of transcripts per type per timepoint.

```{r plot isoform quantification total}
# Plot
ggplot(dt_m_aggr, aes(x = type, y = count, group = type, fill = timepoint)) +
	geom_col(position = "dodge") +
	facet_wrap(~timepoint, nrow = 1) +
	scale_fill_manual(values = timepoints_cols) +
	theme_cowplot() +
	labs(y = "Number of transcripts", x = "") +
	scale_y_continuous(expand = c(0, 0), labels = scales::comma_format()) +
	theme(legend.position = "none",
				text = element_text(family = "Helvetica"),
				axis.text.x = element_text(angle = 45, hjust = 1))
```

Then we plot the number of unique transcripts per type per timepoint.

```{r plot isoform quantification unique}
# Plot Number of unique transcripts
dt_counts_unique_D0 = counts_d0[sum > 0, .N, by = type]
dt_counts_unique_Q1 = counts_q1[sum > 0, .N, by = type]
dt_counts_unique_Q2 = counts_q2[sum > 0, .N, by = type]
dt_counts_unique_Q3 = counts_q3[sum > 0, .N, by = type]

# Combine
dt_counts_unique_D0[, timepoint := "D0"]
dt_counts_unique_Q1[, timepoint := "Q1"]
dt_counts_unique_Q2[, timepoint := "Q2"]
dt_counts_unique_Q3[, timepoint := "Q3"]

dt_counts_unique = rbind(dt_counts_unique_D0, dt_counts_unique_Q1, dt_counts_unique_Q2, dt_counts_unique_Q3)
dt_counts_unique[, type := factor(type, levels = c("known", "nnic", "nic"))]

# Plot
ggplot(dt_counts_unique, aes(x = type, y = N, group = type, fill = timepoint)) +
	geom_col(position = "dodge") +
	facet_wrap(~timepoint, nrow = 1) +
	scale_fill_manual(values = timepoints_cols) +
	theme_cowplot() +
	labs(y = "Number of transcripts", x = "") +
	scale_y_continuous(expand = c(0, 0), labels = scales::comma_format()) +
	theme(legend.position = "none",
				text = element_text(family = "Helvetica"),
				axis.text.x = element_text(angle = 45, hjust = 1))
```